---
title: "Term Project"
author: "Zimu Lyu, 6000-level"
output: html_document
---

# Import Data

```{r, message=FALSE, error=FALSE}
library(tidyverse)
library(zoo) # transform date format
library(lubridate) # extract date feature
library(fastDummies) # for one-hot encoding
library(ggpubr) # multigraph
library(DMwR) # handle unbalanced sample
library(caret) # calculate correlation coefficient
library(corrplot) # visualize correlation matrix
library(caTools) # split data
library(rpart) # decision tree
library(rpart.plot) # visualize tree
library(party) # decision tree
library(randomForest) # random forest
library(car) # calculate vif
library(rawKS) # calculate ks
library(e1071) # svm model
```

```{r, warning=FALSE, message=FALSE}

setwd("/Users/cleardog/BA/MGMT 6962 Data Analytics/Project/Data/")

# import origination data
origclass <- c(
'integer','integer','character','integer','character','real', 'integer','character','real',
'integer','integer','integer','real','character','character','character','character', 'character',
'character','character','character','integer','integer','character','character','character', 'character')
orignames <- c(
'fico','dt_first_pi','flag_fthb','dt_matr','cd_msa','mi_pct','cnt_units','occpy_sts','cltv','dti',
'orig_upb','ltv','int_rt','channel','ppmt_pnlty','prod_type','st','prop_type','zipcode','id_loan',
'loan_purpose','orig_loan_term','cnt_borr','seller_name','servicer_name','flag_sc', 'program_ind'
)

q1_2018_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q12018",".txt"), full.names=F)
q2_2018_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q22018",".txt"), full.names=F)
q3_2018_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q32018",".txt"), full.names=F)
q4_2018_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q42018",".txt"), full.names=F)

readfile_orig <- function(filenm){read.table(filenm,sep="|", header=FALSE, colClasses=origclass)}
q1_2018_orig <- do.call("rbind", lapply(q1_2018_orig,readfile_orig))
q2_2018_orig <- do.call("rbind", lapply(q2_2018_orig,readfile_orig))
q3_2018_orig <- do.call("rbind", lapply(q3_2018_orig,readfile_orig))
q4_2018_orig <- do.call("rbind", lapply(q4_2018_orig,readfile_orig))

names(q1_2018_orig) <- orignames
names(q2_2018_orig) <- orignames
names(q3_2018_orig) <- orignames
names(q4_2018_orig) <- orignames

# import performance data
svcgclass <- c(
'character','integer','real','character','integer','integer','character','character','character',
'integer','real','real','integer','integer','character','integer','integer','integer','integer',
'integer','integer','real','real','character','character','real','real','real', 'character', 'character')
svcgnames <- c(
'id_loan','svcg_cycle','current_upb','delq_sts','loan_age','mths_remng','repch_flag','flag_mod','cd_zero_bal',
'dt_zero_bal','current_int_rt','non_int_brng_upb','dt_lst_pi','mi_recoveries','net_sale_proceeds',
'non_mi_recoveries','expenses','legal_costs','maint_pres_costs','taxes_ins_costs','misc_costs','actual_loss',
'modcost','stepmod_ind','dpm_ind','eltv','zb_removal_upb','dlq_acrd_int',
'disaster_hardship_ind','borrower_assist_flag')

q1_2018_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q12018",".txt"), full.names=F)
q2_2018_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q22018",".txt"), full.names=F)
q3_2018_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q32018",".txt"), full.names=F)
q4_2018_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q42018",".txt"), full.names=F)

readfile_svcg <- function(filenm){read.table(filenm,sep="|", header=FALSE, colClasses=svcgclass)}
q1_2018_svcg <- do.call("rbind", lapply(q1_2018_svcg,readfile_svcg))
q2_2018_svcg <- do.call("rbind", lapply(q2_2018_svcg,readfile_svcg))
q3_2018_svcg <- do.call("rbind", lapply(q3_2018_svcg,readfile_svcg))
q4_2018_svcg <- do.call("rbind", lapply(q4_2018_svcg,readfile_svcg))

names(q1_2018_svcg) <- svcgnames
names(q2_2018_svcg) <- svcgnames
names(q3_2018_svcg) <- svcgnames
names(q4_2018_svcg) <- svcgnames

```

```{r}
orig <- rbind(q1_2018_orig, q2_2018_orig, q3_2018_orig, q4_2018_orig)
svcg <- rbind(q1_2018_svcg, q2_2018_svcg, q3_2018_svcg, q4_2018_svcg)
delq <- svcg[, which(colnames(svcg) %in% c('id_loan','delq_sts'))]
```

Because we need this model to predict the probability of delinquency based on original information, 
so variables in the performance file other than the delinquency status are not used.

```{r}
rm(q1_2018_orig)
rm(q2_2018_orig)
rm(q3_2018_orig)
rm(q4_2018_orig)
rm(q1_2018_svcg)
rm(q2_2018_svcg)
rm(q3_2018_svcg)
rm(q4_2018_svcg)
rm(svcg)
```

```{r}
# merge data
set.seed(100)
loans <- orig %>%
  left_join(delq, by = "id_loan") %>%
  sample_frac(0.01, replace = FALSE)
```

# Data Orverview

Filter single family loans with 30-year fixed term.

```{r}
loans <- filter(loans, orig_loan_term==360 & prop_type=='SF' & prod_type=='FRM')
loans <- loans[,!colnames(loans) %in% c("orig_loan_term","prop_type", "prod_type")]
```

```{r}
head(loans)
str(loans)
```

```{r}
# transform integer format to date format
loans$dt_first_pi <- as.Date(as.yearmon(as.character(loans$dt_first_pi), "%Y%m"))
loans$dt_matr <- as.Date(as.yearmon(as.character(loans$dt_matr), "%Y%m"))
```

```{r}
summary(loans)
```

Many variables have outliers.

```{r}
check_na <- as.data.frame(sapply(loans, function(x) sum(is.na(x))))
colnames(check_na) <- c("count")
arrange(check_na, desc(check_na$count))
```

No missing value. 

```{r error=FALSE, message=FALSE, warning=FALSE}
loans_bydate <- loans %>%
    group_by(dt_first_pi) %>%
    count() 
ggplot(loans_bydate, aes(dt_first_pi, n)) + 
  geom_histogram(stat = 'identity') +
  labs(x = 'First payment date', y = 'count')
```
Most of the data comes from March 2018 to February 2019.     

```{r}
loans <- filter(loans, dt_first_pi>='2018-03-01', dt_first_pi<'2019-03-01')
```

# EDA

Let's first look at the target variable.

## Target variable - Current loan delinquency status

The 90–day delinquency rate is a measure of serious delinquencies. 
It captures borrowers that have missed three or more payments.

Definition from: 
https://www.consumerfinance.gov/data-research/mortgage-performance-trends/mortgages-90-or-more-days-delinquent/

```{r}
table(loans$delq_sts)
```

```{r}
# Delete "R", don't discuss this specific case
loans <- loans[loans$delq_sts!='R',]
# Create a delinquency flag, >3 means >=90
loans$delq <- as.numeric(as.numeric(loans$delq_sts) > 2)
```

Look at delinquency by date features.

```{r, message=FALSE}
monthly_delq <- loans %>%
  group_by(dt_first_pi) %>%
  summarise(count=n(), avgdelq=mean(delq))
monthly_delq

ggplot(monthly_delq, aes(dt_first_pi, avgdelq)) +
  geom_line() +
  labs(x='First Payment Date', y='The Ratio of Delinquency')
```

EDA for some important numeric variables.

## Credit Score(FICO)

```{r}
summary(loans$fico)
# 9999 = Not Available, if Credit Score is < 301 or > 850.
sum(loans$fico==9999)
# Only a few hundred rows columns, so delete them.
loans <- loans[!loans$fico==9999,]
summary(loans$fico)
```

```{r, message=FALSE}
loans %>%
  group_by(dt_first_pi) %>%
  summarise(count=n(), avgfico=mean(fico))

loans %>%
  group_by(dt_first_pi) %>%
  summarise(count=n(), avgfico=mean(fico)) %>%
  ggplot() +
  geom_line(aes(dt_first_pi, avgfico)) +
  labs(x='First Payment Date', y='Mean Credit Score')
```
Relationship with delinquency

```{r}
delq_fico <- loans %>% select(delq_sts, fico) %>% mutate(delq_sts=as.integer(delq_sts))
qplot(factor(delq_sts), fico, data=delq_fico, geom='boxplot') + labs(x='Delinquency Status', y='FICO')
```

## Mortgage insurance percentage(MI%)

```{r}
summary(loans$mi_pct)
table(factor(loans$mi_pct))
mean(loans$mi_pct==0) # 0 = no MI, 66% of the value is zero
loans <- loans[!loans$mi_pct==999,] # 999 = not available(<1 or >55), delete them
```

```{r, message=FALSE}
mi_delq <- loans %>%
  mutate(mi_pct=factor(mi_pct)) %>%
  group_by(mi_pct) %>%
  summarise(delq_pct = mean(delq))
mi_delq

have_mi_delq <- loans %>%
  mutate(mi=as.integer(mi_pct==0)) %>%
  group_by(mi) %>%
  summarise(delq_pct = mean(delq))
have_mi_delq
```

```{r}
mi1 <- qplot(loans$mi_pct, binwidth=1, xlab = 'Distribution of MI%', ylab = 'Count')
mi2 <- ggplot(mi_delq, aes(mi_pct, delq_pct)) + 
  geom_bar(stat = "identity") +
  labs(x = 'MI%', y = 'Delinquency Ratio')
mi3 <- ggplot(have_mi_delq, aes(factor(mi), delq_pct)) + geom_bar(stat = "identity") + 
  labs(x = 'MI%', y = 'Delinquency Ratio')

ggarrange(mi1, mi2, mi3, ncol = 3, nrow = 2, widths = c(2,3,2)) 
```

## Original combined loan-to-value(CLTV)

```{r}
summary(loans$cltv)
sum(loans$cltv==999)
# 999: cltv<80 or cltv>200, or cltv<ltv
loans <- loans[!loans$cltv==999,]
```

## Original debt-to-income(DTI)

```{r}
summary(loans$dti)
sum(loans$dti==999)
# 999: dti>65
loans <- loans[!loans$dti==999,]
```

## Original unpaid principal balance 

```{r}
summary(loans$orig_upb)
```

## Original loan-to-value(LTV)

```{r}
summary(loans$ltv)
sum(loans$ltv==999)
# 999: ltv<6 or ltv>105
loans <- loans[!loans$ltv==999,]
```

## Original interest rate

```{r}
summary(loans$int_rt)
```

## Histograms of Numeric Variables

```{r, message=FALSE}
fig1 <- qplot(loans$fico, xlab = 'Distribution of Credit Score')
fig2 <- qplot(loans$cltv, xlab = 'Distribution of CLTV')
fig3 <- qplot(loans$dti, xlab = 'Distribution of DTI')
fig4 <- qplot(loans$orig_upb, xlab = 'Distribution of Original UPB')
fig5 <- qplot(loans$ltv, xlab = 'Distribution of LTV')
fig6 <- qplot(loans$int_rt, xlab = 'Distribution of Interest Rate')

ggarrange(fig1, fig2, fig3, fig4, fig5, fig6, ncol = 3, nrow = 2) 
```

## Box plots 

```{r}
fig1 <- qplot(factor(delq), fico, data=loans, geom='boxplot') + labs(x='Delinquency', y='FICO')
fig2 <- qplot(x=factor(delq), y=cltv, data=loans, geom='boxplot', xlab = 'Delinquency', ylab = 'CLTV')
fig3 <- qplot(x=factor(delq), y=dti, data=loans, geom='boxplot', xlab = 'Delinquency', ylab = 'DTI') 
fig4 <- qplot(x=factor(delq), y=orig_upb, data=loans, geom='boxplot', xlab = 'Delinquency', ylab = 'Original UPB') 
fig5 <- qplot(x=factor(delq), y=ltv, data=loans, geom='boxplot', xlab = 'Delinquency', ylab = 'LTV') 
fig6 <- qplot(x=factor(delq), y=int_rt, data=loans, geom='boxplot', xlab = 'Delinquency', ylab = 'Interest Rate') 

ggarrange(fig1, fig2, fig3, fig4, fig5, fig6, ncol = 3, nrow = 2) 
```

Next, we'll explore categorical variables.

## First time homebuyer flag

```{r, message=FALSE}
sum(loans$flag_fthb=='')
table(loans$flag_fthb)
# '9' should be considered
```

```{r, message=FALSE}
delq_fthb <- loans %>%
  group_by(flag_fthb) %>%
  summarise(delq_pct=mean(delq))
delq_fthb
```

## Metropolitan statistical area or metropolitan division

```{r}
length(unique(loans$cd_msa))
sum(loans$cd_msa=='')
mean(loans$cd_msa=='')
loans[loans$cd_msa=='',]$cd_msa = 0 # transform space to 0
```

```{r, message=FALSE}
# different area
metro_count <- loans %>%
    group_by(cd_msa) %>%
    count() %>%
    arrange(desc(n)) 
head(metro_count)

delq_metro <- loans %>%
    group_by(cd_msa) %>%
    summarize(delq_pct=mean(delq)) %>%
    arrange(desc(delq_pct))
head(delq_metro)

# metro vs no-metro
delq_is_metro <- loans %>%
    mutate(is_metro = (loans$cd_msa!='0')) %>%
    group_by(is_metro) %>%
    summarize(delq_pct=mean(delq))
delq_is_metro
```

## Number of units

```{r, message=FALSE}
table(loans$cnt_units)
# should use one-hot encoding

delq_unit <- loans %>%
  group_by(cnt_units) %>%
  summarise(delq_pct=mean(delq))
delq_unit
```

## Occupancy status

```{r, message=FALSE}
table(loans$occpy_sts)
# P = Primary residence, I = Investment property, S = Second home
# one-hot

delq_occpy <- loans %>%
  group_by(occpy_sts) %>%
  summarise(delq_pct=mean(delq))
delq_occpy
```

## Channel

```{r, message=FALSE}
table(loans$channel)
# Broker / Correspondent / Retail / TPO not specified
# one-hot

# T should be deleted
loans <- loans[!loans$channel=='T',]
```

```{r, message=FALSE}
delq_chnl <- loans %>%
  group_by(channel) %>%
  summarise(delq_pct=mean(delq))
delq_chnl
```

## Prepayment penalty mortgage(PPM) flag

```{r, message=FALSE}
table(loans$ppmt_pnlty) # Space means Y = PPM.
loans[loans$ppmt_pnlty=='',]$ppmt_pnlty = "Y"
# one-hot
```

```{r,message=FALSE}
delq_ppm <- loans %>%
  group_by(ppmt_pnlty) %>%
  summarise(delq_pct=mean(delq))
delq_ppm
```

## Loan purpose

```{r, message=FALSE}
table(loans$loan_purpose)
# Cash-out refinance / No cash-out refinance / Purchase
# one-hot

delq_purpose <- loans %>%
  group_by(loan_purpose) %>%
  summarise(delq_pct=mean(delq))
delq_purpose
```

## Number of borrowers

```{r, message=FALSE}
table(factor(loans$cnt_borr))

delq_borr <- loans %>%
  group_by(cnt_borr) %>%
  summarise(delq_pct=mean(delq))
delq_borr
```

## Super conforming flag

```{r, message=FALSE}
table(loans$flag_sc)
# Space = Not super conforming
loans[loans$flag_sc=='',]$flag_sc = "N"
```

```{r, message=FALSE}
delq_sc <- loans %>%
  group_by(flag_sc) %>%
  summarise(delq_pct=mean(delq))
delq_sc
```

## Bar plots

```{r}
fig1 <- ggplot(delq_fthb,aes(flag_fthb, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'First Time Homebuyer Flag', y = 'Delinquency Ratio')
fig2 <- ggplot(delq_is_metro,aes(is_metro, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Metropolitan', y = 'Delinquency Ratio')
fig3 <- ggplot(delq_unit,aes(cnt_units, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Number of Units', y = 'Delinquency Ratio')
fig4 <- ggplot(delq_occpy,aes(occpy_sts, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Occupancy Status', y = 'Delinquency Ratio')
fig5 <- ggplot(delq_chnl,aes(channel, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Channel', y = 'Delinquency Ratio')
fig6 <- ggplot(delq_ppm,aes(ppmt_pnlty, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Prepayment Penalty Mortgage Flag', y = 'Delinquency Ratio')
fig7 <- ggplot(delq_purpose,aes(loan_purpose, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Loan Purpose', y = 'Delinquency Ratio')
fig8 <- ggplot(delq_borr,aes(factor(cnt_borr), delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Number of Borrowers', y = 'Delinquency Ratio')
fig9 <- ggplot(delq_sc,aes(flag_sc, delq_pct)) + geom_bar(stat = "identity") +
  labs(x = 'Super Conforming Flag', y = 'Delinquency Ratio')

ggarrange(fig1, fig2, fig3, fig4, fig5, fig6, ncol = 3, nrow = 2)
```
```{r}
ggarrange(fig7, fig8, fig9, ncol = 3, nrow = 2)
```

## Property state

```{r}
table(loans$st)
# Nation and state? EDA
# one-hot
```

```{r, message=FALSE}
loans %>%
  filter(st == 'CA') %>%
  group_by(dt_first_pi) %>%
  summarise(statedelq=mean(delq)) %>%
  ggplot(aes(dt_first_pi, statedelq)) +
  geom_line()

loans %>%
  filter(st == 'NY') %>%
  group_by(dt_first_pi) %>%
  summarise(statedelq=mean(delq)) %>%
  ggplot(aes(dt_first_pi, statedelq)) +
  geom_line()
```

```{r, message=FALSE}
loans %>%
    group_by(st) %>%
    count() %>%
    arrange(desc(n))

delq_state <- loans %>%
    group_by(st) %>%
    summarize(delq_pct=mean(delq)) %>%
    arrange(desc(delq_pct)) 

head(delq_state, 10)

ggplot(delq_state,aes(reorder(st, delq_pct), delq_pct))+
  geom_bar(stat = "identity") + 
  coord_flip() + 
  labs(x = 'State', y = 'The Ratio of Delinquency') +
  theme(axis.text.y = element_text(size = 6))
```

## Seller name

```{r}
qplot(fct_rev(fct_infreq(loans$seller_name)), geom = "bar", xlab = 'Seller Name') + coord_flip()
```

```{r, message=FALSE}
delq_seller <- loans %>%
    group_by(seller_name) %>%
    summarize(delq_pct=mean(delq)) %>%
    arrange(desc(delq_pct)) 

head(delq_seller, 10)

ggplot(delq_seller,aes(reorder(seller_name,delq_pct), delq_pct))+
  geom_bar(stat = "identity") + 
  labs(x = 'Seller Name', y = 'The Ratio of Delinquency') +
  coord_flip()
```

## Servicer name

```{r}
qplot(fct_rev(fct_infreq(loans$servicer_name)), geom = "bar", xlab = 'Servicer Name') + coord_flip()
```

```{r, message=FALSE}
delq_servicer <- loans %>%
    group_by(servicer_name) %>%
    summarize(delq_pct=mean(delq)) %>%
    arrange(desc(delq_pct)) 

head(delq_servicer, 10)

ggplot(delq_servicer,aes(reorder(servicer_name,delq_pct), delq_pct))+
  geom_bar(stat = "identity") + 
  labs(x = 'Servicer Name', y = 'The Ratio of Delinquency') +
  coord_flip()
```

## Date Feature

```{r}
loans$dt_month <- as.character(month(loans$dt_first_pi))
```

# Analysis and Feature Engineering

## Create a correlation matrix for numeric variables

```{r}
num_vars <- c("delq","fico","mi_pct","cnt_units","cltv","dti","orig_upb","ltv","int_rt","cnt_borr")
correlationMatrix <- cor(loans[num_vars])
print(correlationMatrix)
```

```{r}
corrplot.mixed(correlationMatrix, lower = "color", upper = "number", tl.pos = "lt", diag = "l", tl.col = "black")
```

```{r}
qplot(ltv, cltv, data = loans, xlab = 'LTV', ylab = 'CLTV')
qplot(ltv, mi_pct, data = loans, xlab = 'LTV', ylab = 'MI%')
```

LTV should be deleted.

```{r}
# delete useless columns
loans <- loans[,!colnames(loans) %in% c("dt_first_pi","dt_matr","zipcode","id_loan","program_ind","delq_sts", "ltv")]
```

## Create new data using rose/smote

```{r}
loans <- loans[!loans$st=='PR',] 
# very little data from this state and no delinquency record.
# if create dummy variable for it, all rows will be 0. It makes no sense.

cat_vars <- c("flag_fthb","cd_msa","occpy_sts","channel","ppmt_pnlty","st","loan_purpose",
              "seller_name","servicer_name","flag_sc","dt_month","delq")
loans[,cat_vars] <- lapply(loans[,cat_vars], as.factor)

table(loans$delq)
set.seed(233)
newdata <- SMOTE(delq ~ ., data = loans, perc.over = 43000, perc.under=100) ###
table(newdata$delq)
```

## Create dummy variables for logistic regression

```{r}
# transform to a binary variable
newdata$mi_pct <- as.integer(newdata$mi_pct > 0) #??
newdata$cd_msa <- as.integer(newdata$cd_msa!='0')
newdata$ppmt_pnlty <- as.integer(newdata$ppmt_pnlty=='Y')
newdata$flag_sc <- as.integer(newdata$flag_sc == 'Y')

# create dummy variable
cat_vars <- c("flag_fthb","occpy_sts","channel","loan_purpose","seller_name", "servicer_name", "dt_month")
newdata <- dummy_cols(newdata, select_columns = cat_vars, remove_first_dummy = TRUE, remove_selected_columns = TRUE)
newdata <- dummy_cols(newdata, select_columns = "st", remove_first_dummy = FALSE, remove_selected_columns = TRUE)

# convert into factors
dummy_vars <- names(newdata[,13:139])
cat_vars <- c("mi_pct","cd_msa","ppmt_pnlty","flag_sc", dummy_vars)
newdata[,cat_vars] <- lapply(newdata[,cat_vars], as.factor)
```

# Modeling

```{r}
# split data into train and test sets
set.seed(233)
split = sample.split(newdata$delq, SplitRatio = .8)
train = subset(newdata, split == TRUE)
test  = subset(newdata, split == FALSE)
# delete space in column names
names(train) <- make.names(names(train)) 
names(test) <- make.names(names(test))
```

## Logistic Regression

```{r}
set.seed(233)
logit <- glm(delq ~ ., data = train, family = "binomial")
# summary(logit)
```

```{r}
set.seed(233)
prob <- predict(logit, test)
logit_pred <- factor(prob > .5, levels = c(FALSE, TRUE), labels = c(0, 1))
logit_pref <- table(test$delq, logit_pred, dnn = c("Actual", "Predicted"))
logit_pref
```

```{r}
ks_table(as.numeric(logit_pred),test[,12]) %>% summarise(ks = max(ks))
```

```{r}
# VIF for numeric variables
num_vars <- c("delq","fico","mi_pct","cnt_units","cltv","dti","orig_upb","int_rt","cnt_borr")
train2 <- train[num_vars]
vif(glm(delq ~ ., data = train2, family = "binomial"))
```

```{r, results='hide'}
logit_reduced <- step(logit, trace=0, steps=200)
```

```{r}
# summary(logit_reduced)
```

```{r}
prob <- predict(logit_reduced, test)
logit_reduced_pred <- factor(prob > .5, levels = c(FALSE, TRUE), labels = c(0, 1))
logit_reduced_pref <- table(test$delq, logit_reduced_pred, dnn = c("Actual", "Predicted"))
logit_reduced_pref
```

```{r}
performance <- function(table, n=2) {
  tn = table[1,1]
  fp = table[1,2]
  fn = table[2,1]
  tp = table[2,2]
  sensitivity = tp/(tp+fn)
  specificity = tn/(tn+fp)
  ppp = tp/(tp+fp)
  npp = tn/(tn+fn)
  accuracy = (tp+tn)/(tp+tn+fp+fn)
  result <- paste("Sensitivity = ", round(sensitivity, n),
                  "\nSpecificity = ", round(specificity, n),
                  "\nPositive Predictive Value = ", round(ppp, n),
                  "\nNegative Predictive Value = ", round(npp, n),
                  "\nAccuracy = ", round(accuracy, n), "\n", sep = "")
  cat(result)
}
```

```{r}
performance(logit_pref)
performance(logit_reduced_pref)
```

```{r}
ks_table(as.numeric(logit_reduced_pred),test[,12]) %>% summarise(ks = max(ks))
```

## Decision Tree

### Use rpart
```{r}
set.seed(233)
tree <- rpart(delq ~ ., data = train, method = 'class')
tree$cptable
plotcp(tree)
```

```{r}
rpart.plot(tree)
tree_pruned <- prune(tree, cp=.01066713)
rpart.plot(tree_pruned) 
```

```{r}
tree_pred <- predict(tree_pruned, test[,-12], type = "class")
tree_pref <- table(test[,12], tree_pred, dnn = c("Actual", "Predicted"))
tree_pref
```

```{r}
performance(tree_pref)
```
```{r}
ks_table(as.numeric(tree_pred),test[,12]) %>% summarise(ks = max(ks))
```

### Use ctree

```{r}
set.seed(233)
ctree <- ctree(delq ~ ., data = train)
```

```{r}
ctree_pred <- predict(ctree, test[,-12])
ctree_pref <- table(test[,12], ctree_pred, dnn = c("Actual", "Predicted"))
ctree_pref
```

```{r}
performance(ctree_pref)
```

```{r}
ks_table(as.numeric(ctree_pred),test[,12]) %>% summarise(ks = max(ks))
```

## Random Forest

```{r}
set.seed(233)
forest <- randomForest(delq ~ ., data = train, importance = TRUE)
forest
```

```{r, out.width='70%', fig.align='center'}
# importance(forest, type = 2)
varImpPlot(forest, type = 2, n.var = 20)
```

```{r}
forest_pred <- predict(forest, test, type = "class")
forest_pref <- table(test$delq, forest_pred, dnn = c("Actual", "Predicted"))
forest_pref
```

```{r}
ks_table(as.numeric(forest_pred),test[,12]) %>% summarise(ks = max(ks))
```

```{r}
performance(forest_pref, n=3)
```

## SVM

```{r}
svm <- svm(delq ~ ., data = train) # very slow
svm
```

```{r}
set.seed(233)
svm_pred <- predict(svm, test)
svm_pref <- table(test$delq, svm_pred, dnn = c("Actual", "Predicted"))
svm_pref
```

```{r}
performance(svm_pref,3)
```

```{r}
ks_table(as.numeric(svm_pred),test[,12]) %>% summarise(ks = max(ks))
```

```{r}
# set.seed(233)
# tuned <- tune.svm(delq ~ ., data = train, gamma = 10^(-2:2), cost = 10^(-2:2))
# tuned 
## too slow and Rstudio always breaks when running these codes. My computer cannot do it
```

# Import and preprocess 2019 data (test set)

```{r, warning=FALSE, message=FALSE}
setwd("/Users/cleardog/BA/MGMT 6962 Data Analytics/Project/Data/")

q1_2019_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q12019",".txt"), full.names=F)
q2_2019_orig <- list.files(path=getwd(), pattern=c("historical_data1_Q22019",".txt"), full.names=F)
readfile_orig <- function(filenm){read.table(filenm,sep="|", header=FALSE, colClasses=origclass)}
q1_2019_orig <- do.call("rbind", lapply(q1_2019_orig,readfile_orig))
q2_2019_orig <- do.call("rbind", lapply(q2_2019_orig,readfile_orig))
names(q1_2019_orig) <- orignames
names(q2_2019_orig) <- orignames

q1_2019_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q12019",".txt"), full.names=F)
q2_2019_svcg <- list.files(path=getwd(), pattern=c("historical_data1_time_Q22019",".txt"), full.names=F)
readfile_svcg <- function(filenm){read.table(filenm,sep="|", header=FALSE, colClasses=svcgclass)}
q1_2019_svcg <- do.call("rbind", lapply(q1_2019_svcg,readfile_svcg))
q2_2019_svcg <- do.call("rbind", lapply(q2_2019_svcg,readfile_svcg))
names(q1_2019_svcg) <- svcgnames
names(q2_2019_svcg) <- svcgnames

orig2 <- rbind(q1_2019_orig, q2_2019_orig)
svcg2 <- rbind(q1_2019_svcg, q2_2019_svcg)
delq2 <- svcg2[, which(colnames(svcg2) %in% c('id_loan','delq_sts'))]
rm(q1_2019_orig)
rm(q2_2019_orig)
rm(q1_2019_svcg)
rm(q2_2019_svcg)
rm(svcg2)
```

```{r}
# table(as.numeric(as.numeric(delq2$delq_sts) > 2)) 
# 0: 5107212 1: 1626
set.seed(100)
all_loans <- orig2 %>%
  left_join(delq2, by = "id_loan")

# select about 1000+ loans not in delinquency  
loans_0 <- all_loans %>%
  filter(as.numeric(delq_sts) <= 2) %>%
  sample_frac(0.0002, replace = FALSE)
loans_1 <- all_loans%>% filter(as.numeric(delq_sts) > 2)
loans_test <- rbind(loans_0, loans_1)

loans_test <- filter(loans_test, orig_loan_term==360 & prop_type=='SF' & prod_type=='FRM')
loans_test <- loans_test[,!colnames(loans_test) %in% c("orig_loan_term","prop_type", "prod_type","dt_matr","zipcode","id_loan","program_ind", "ltv")]
loans_test$dt_first_pi <- as.Date(as.yearmon(as.character(loans_test$dt_first_pi), "%Y%m"))

# check_na <- as.data.frame(sapply(loans_test, function(x) sum(is.na(x))))
# colnames(check_na) <- c("count")
# arrange(check_na, desc(check_na$count))
# No null value

loans_test <- filter(loans_test, dt_first_pi>='2019-03-01', dt_first_pi<='2019-08-01')
loans_test <- loans_test[loans_test$delq_sts!='R',]
loans_test$delq <- as.numeric(as.numeric(loans_test$delq_sts) > 2)
loans_test <- loans_test[!loans_test$fico==9999,]
loans_test <- loans_test[!loans_test$mi_pct==999,]
loans_test <- loans_test[!loans_test$cltv==999,]
loans_test <- loans_test[!loans_test$dti==999,]
loans_test[loans_test$cd_msa=='',]$cd_msa = 0
# loans_test <- loans_test[!loans_test$channel=='T',] 
loans_test[loans_test$ppmt_pnlty=='',]$ppmt_pnlty = "Y"
loans_test[loans_test$flag_sc=='',]$flag_sc = "N"
loans_test$dt_month <- as.character(month(loans_test$dt_first_pi))
loans_test <- loans_test[,!colnames(loans_test) %in% c("delq_sts","dt_first_pi")]
loans_test <- loans_test[!loans_test$st=='PR',] 

cat_vars <- c("flag_fthb","cd_msa","occpy_sts","channel","ppmt_pnlty","st","loan_purpose",
              "seller_name","servicer_name","flag_sc","dt_month","delq")
#loans[,cat_vars] <- lapply(loans[,cat_vars], as.factor)
loans_test$mi_pct <- as.integer(loans_test$mi_pct > 0)
loans_test$cd_msa <- as.integer(loans_test$cd_msa!='0')
loans_test$ppmt_pnlty <- as.integer(loans_test$ppmt_pnlty=='Y')
loans_test$flag_sc <- as.integer(loans_test$flag_sc == 'Y')
cat_vars <- c("flag_fthb","occpy_sts","channel","loan_purpose")
loans_test <- dummy_cols(loans_test, select_columns = cat_vars, remove_first_dummy = TRUE, 
                         remove_selected_columns = TRUE)

# for seller name, servicer name and month
seller_cols <- unique(loans$seller_name)[-1]
servicer_cols <- unique(loans$servicer_name)[-1]
month_cols <- unique(loans$dt_month)[-4]
st_cols <- unique(loans$st)
for (name in seller_cols) {
  for (i in 1:dim(loans_test)[1]) {
    if (loans_test[['seller_name']][i] == name) {
      loans_test[i, str_c('seller_name_', name)] = 1
    } else {
      loans_test[i, str_c('seller_name_', name)] = 0
    }
  }
}
loans_test['seller_name_Other sellers'] = 1-rowSums(loans_test[, 25:53])
for (name in servicer_cols) {
  for (i in 1:dim(loans_test)[1]) {
    if (loans_test[['servicer_name']][i] == name) {
      loans_test[i, str_c('servicer_name_', name)] = 1
    } else {
      loans_test[i, str_c('servicer_name_', name)] = 0
    }
  }
}
loans_test['servicer_name_Other servicers'] = 1-rowSums(loans_test[, 55:80])
for (name in month_cols) {
  for (i in 1:dim(loans_test)[1]) {
    if (loans_test[['dt_month']][i] == name) {
      loans_test[i, str_c('dt_month_', name)] = 1
    } else {
      loans_test[i, str_c('dt_month_', name)] = 0
    }
  }
}
for (name in st_cols) {
  for (i in 1:dim(loans_test)[1]) {
    if (loans_test[['st']][i] == name) {
      loans_test[i, str_c('st_', name)] = 1
    } else {
      loans_test[i, str_c('st_', name)] = 0
    }
  }
}
loans_test <- loans_test[,!colnames(loans_test) %in% c("dt_month","seller_name","servicer_name", "st",
              "seller_name_AMERIHOME MORTGAGE COMPANY, LLC", "servicer_name_AMERIHOME MORTGAGE COMPANY, LLC")]
names(loans_test) <- make.names(names(loans_test)) 

# classes of new data should match original data (for ctree)
vars1 <- c("fico","cnt_units","dti","orig_upb","cnt_borr")
vars2 <- names(loans_test)[21:139]
loans_test[,vars1] <- lapply(loans_test[,vars1], as.numeric)
loans_test[,vars2] <- lapply(loans_test[,vars2], as.integer)
```

# Prediction on the test set

```{r}
prob <- predict(logit_reduced, loans_test)
logit_reduced_pred <- factor(prob > .5, levels = c(FALSE, TRUE), labels = c(0, 1))
logit_reduced_pref <- table(loans_test$delq, logit_reduced_pred, dnn = c("Actual", "Predicted"))
logit_reduced_pref
performance(logit_reduced_pref)
```

```{r}
ks_table(as.numeric(logit_reduced_pred),loans_test[,12]) %>% summarise(ks = max(ks))
```

```{r}
tree_pred <- predict(tree_pruned, loans_test[,-12], type = "class")
tree_pref <- table(loans_test$delq, tree_pred, dnn = c("Actual", "Predicted"))
tree_pref
performance(tree_pref)
```

```{r}
ks_table(as.numeric(tree_pred),loans_test[,12]) %>% summarise(ks = max(ks))
```

```{r}
ctree_pred <- predict(ctree, loans_test[,-12])
ctree_pref <- table(loans_test$delq, ctree_pred, dnn = c("Actual", "Predicted"))
ctree_pref
performance(ctree_pref)
```

```{r}
ks_table(as.numeric(ctree_pred),loans_test[,12]) %>% summarise(ks = max(ks))
```

```{r}
forest_pred <- predict(forest, loans_test, type = "class")
forest_pref <- table(loans_test$delq, forest_pred, dnn = c("Actual", "Predicted"))
forest_pref
performance(forest_pref)
```

```{r}
ks_table(as.numeric(forest_pred),loans_test[,12]) %>% summarise(ks = max(ks))
```

```{r}
svm_pred <- predict(svm, loans_test)
svm_pref <- table(loans_test$delq, svm_pred, dnn = c("Actual", "Predicted"))
svm_pref
performance(svm_pref)
```

```{r}
ks_table(as.numeric(svm_pred),loans_test[,12]) %>% summarise(ks = max(ks))
```

